---
title: "UK Road Traffic Accident Analysis (2024)"
subtitle: "Exploration, Preparation, and Risk Factor Analysis"
author: "Tiago De Oliveira Freitas - 2021406"
date: "November 30, 2025"
output: 
  html_document
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: cosmo
    highlight: tango
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, 
                      fig.width = 10, fig.height = 6, dpi = 150)
```
# 1. Introduction and Project Goal {.tabset}

## 1.1. Context

This project conducts a comprehensive exploratory analysis of road traffic accident data in the United Kingdom for the year 2024, utilizing public data from the **UK Department for Transport**.

## 1.2. Research Question

> **What factors are most correlated with severe traffic accidents?**

This question is fundamental for:

- Identifying key risk factors
- Informing public road safety policies
- Reducing serious accidents and fatalities

## 1.3. Methodology

The project follows these steps:

1.  **Initial Exploration**: Load and understand the data structure.
2.  **Cleaning and Preparation**: Handle missing values and create new features.
3.  **Exploratory Data Analysis (EDA)**: Visualize relationships and patterns.
4.  **Encoding and PCA**: Prepare data for modeling and reduce dimensionality.
5.  **Conclusion**: Present findings and their implications.

---

# 2. Loading Libraries and Data

## 2.1. Installing and Loading Packages

```{r libraries}
# Install and load necessary packages
required_packages <- c("tidyverse", "ggplot2", "corrplot", "psych", 
                       "gridExtra", "scales", "RColorBrewer", 
                       "FactoMineR", "factoextra", "knitr", "kableExtra")

for(pkg in required_packages) {
  if(!require(pkg, character.only = TRUE, quietly = TRUE)) {
    install.packages(pkg, repos = "https://cloud.r-project.org/", quiet = TRUE)
    library(pkg, character.only = TRUE)
  }
}

cat("All packages loaded successfully!\n")
```